Follow-up Samples

### 2017-03-15 submission/manuscript_outline_20161024.Rmd

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: reference.bib
---

@@

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had no difference in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that the lesion associated community persists after treatment.  A treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that there are members within the community that repond to treatment.  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*).

**Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.  Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.  Commensal bacteria were crucial for accurate model classification, suggesting that these bacteria may be important to initial polyp formation and transition to carcinoma.

@@

Recent studies in humans have shown that both the bacterial community and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].  Further, Dejea, et al. observed that bacterial communities are altered between normal and tumor tissue [@dejea_microbiota_2014].  Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.  Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.  Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].  They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).

In this study, we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].  We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up treatment samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether they were changing after treatment and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify which were important for classifying both lesion versus normal and initial versus follow up treatment samples.  Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionally treatments significantly lowered any of the metrics examined versus those who received only lesion removal.  This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***The Lesion Model:***
We tested whether follow up samples were more normal than initial samples by building a model based solely on OTUs alone that classified lesion versus normal.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100-different test set AUC iterations [Figure 2A].  There was a total of `r length(lesion_top10_varsMDA$variable)` OTUs that were used in this model [Figure 2B].  *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [Figure 2B].  Heavily represented genera within the model included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.

If there was movement towards a more normal community after treatment of lesion, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.  We observed no such decrease in positive probability for either those with adenoma or carcinoma [Figure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  We also observed a significant difference between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  However, this model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].

***The Treatment Model:***
To test if there were differences that could be used to classify initial and follow up treatment samples we built a model to classify initial samples.  This treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better than the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3A].  There was a total of `r length(IF_top10_varsMDA$variable)` OTUs that were used for this model [Figure 3B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].  Similar to the lesion model, heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.

If there were changes between initial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.  The is what we observed for the treatment model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  When we separated lesion into adenoma and carcinoma there was a decrease in positive probability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [Figure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  For this model, there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).

***Common OTUs to both Models:***
We next wanted to know what predictors within the lesion model were also in the initial sample model.  The main purpose was to identify which OTUs could be important in lesion formation and were impacted by treatment.  When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000061")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000037")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000006")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000015")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000167")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  These along withthe vast majority of the OTUs that were common between models had classifications to bacteria typically thought of as commensal [Table S2].

***Chemotherapy and Radiation Differences:*** 
After observing these changes due to treatment we assessed whether chemotherapy or radiation impacted the observed results.  Only the treatment model had a significant change in positive probability for those treated with chemotherapy or radiation versus those who received neither [Figure 4A & Table S3] (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05). Additionally, there was no difference between the change in positive probability for chemotherapy and radiation therapy (Uncorrected P-value `r ifelse(chemo_rad_summary['chemo_v_rads_IF', 'pvalue'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  Using the `r length(common_vars_summary$otu)` common OTUs we tested how chemotherapy or radiation affects important members of the community.  For both chemotherapy and radiation, the directionality of change of these OTUs were similar [Figure 4B & 4C].  After multiple comparison correction, only *`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "lowest_ID"]`* (`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "otu"]`) remained significantly different between those who did and did not receive chemotherapy.  This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to effects on the bacterial community.

@@

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].  Here we describe how the bacterial community changes after treatment and which OTUs may be most important to lesion classification.  Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes.  This suggest that members within the commensal community may be the first that change during CRC pathogenesis.  These changes, in turn, could be the first step in allowing more inflammatory bacteria to gain a foothold within the colon [@flynn_metabolic_2016].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models. There are several reasons why this may have occurred. First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy. Second, is that our Random Forest models were able to gather the same information from other OTUs. Third, is that changes in commensal bacteria are the common differences across adenoma and carcinoma. Finally, it is also possible that all these explanations could have played a role.  Our observations though, would suggest that an individual's resident bacteria have a large role to play in polyp formation and could change in a way that allows predictive models to lower the positive probability of lesion or initial sample before treatment [Figure 2 & 3]. It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after treatment. It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after treatment [Figure S2].

>>>>

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  word_document: default
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: reference.bib
---

@@

**Results:** There was a larger change in the bacterial community in response to lesion removal for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples decreased in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting a movement towards a normal bacterial community.  An initial sample model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with the majority of these classifying to commensal bacteria (e.g. *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000018")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000009")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*).

**Conclusions:** Our data supports the hypothesis that the bacterial community changes after treatment.  Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.  Commensal bacteria were crucial for accurate model classification, suggesting that these bacteria may be important to initial polyp formation and transition to carcinoma.

@@

Recent studies in humans have shown that both the bacterial community and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].  Further, Dejea, et al. observed that bacterial communities are altered between normal and tumor tissue [@dejea_microbiota_2014].  Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.  Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.  Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies show how our gut bacteria can impact CRC progression via a changed community or invations by more inflammatory bacteria [@flynn_metabolic_2016].  They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).

In this study we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].  We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether initial and follow up samples were changing and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify the crucial OTUs involved with both classifying lesion or normal and initial versus follow up samples.  Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionaly treatments significantly lowered any of the metrics examined versus those who received only lesion removal.  This study will provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***The Lesion Model:***
We tested whether follow up samples were more normal than initial samples by building a model that classified lesion versus normal.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold corss validations.  The ROC curve for the final lesion model used was within the observed range of the 100-different test set AUC iterations [Figure 2A].  There was a total of `r length(lesion_top10_varsMDA$variable)` OTUs that were used in this model [Figure 2B].  *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [FIgure 2B].  Heavily represented genera within the model included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.

If there was movement towards a more normal community after treatment of lesion, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.  We observed no such decrease in positive probability for either those with adenoma or carcinoma [FIgure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  We also observed a significant difference between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  This model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].

***The Treatment Model:***
To test if there were differences that could be used to classify initial and follow up treatment samples we built a model to classify initial samples.  This treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better than the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3A].  There was a total of `r length(IF_top10_varsMDA$variable)` OTUs that were used for this model [Figure 3B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].  Similarly, to the lesion model heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.

If there were changes between intial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.  The is what we observed for the treatment model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  When we separated lesion into adenoma and carcinoma there was a decrease in positive arobability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [FIgure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  For this model, there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).

***Common OTUs to both Models:***
We next wanted to know what predictors within the lesion model were also in the initial sample model.  The main purpose was to identify which TOUs could be important in lesion formation and were impacted by treatment.  When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000061")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000037")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000006")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000015")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000167")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  These along withthe vast majority of the OTUs that were common between models had classifications to bacteria typically thought of as commensal [Table S2].

***Chemotherapy and Radiation Differences:*** 
After observing these changes due to treatment we assessed whether chemotherapy or radiation impacted the observed results.  Only the treatment model had a significant change in positive probability for those treated with chemotherapy or radiation versus those who received neither [Figure 4A & Table S3] (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05). Using the `r length(common_vars_summary$otu)` common OTUs we tested how chemotherapy or radiation affects important members of the community.  FOr both chemotherapy or radiation, the directionaly of change of these OTUs were similar [Figure 4B & 4C].  After multiple comparison correction, only *`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "lowest_ID"]`* (`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "otu"]`) remained significantly different between those who did and did not receive chemotherapy.  This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to affects on the bacterial community.

@@

## Discussion
This study builds upon previous work from numerous labs that have considered both how the bacterial community between those with and without CRC differ and how it might be used as an early screening tool [@yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013].  Here we describe how the bacterial community changes after treatment and which OTUs may be most important to lesion classification.  Interestingly, many of the most important OTUs for both treatment and lesion classification had taxonomic identification for resident gut microbes which could suggest that members within the commensal community may be the first that change during pathogenesis of disease.  Further, these changes could be the first step in allowing more inflammatory bacterium to gain a foothold within the colon [@flynn_metabolic_2016].

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models. There are several reasons why this may have occurred. First, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy. Second, is that our Random Forest models were able to gather the same information from other OTUs. Third, it is also possible that changes in commensal bacteria are the common differences across adenoma and carcinoma. Finally, it is also possible that all these explanations could have played a role.  Our observations though, would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of wither lesion or initial sample before treatment [Figure 2 & 3]. It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that they are not the main bacteria changing after treatment. It is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances in carcinoma and not adenoma individuals but also large decreases in relative abundance in some of those with carcinoma but not in those with adenoma after lesion removal [Figure S2].

### 2017-03-15 submission/supplemental_outline_20161024.Rmd

@@

variable <- c(rep("Positive Probability", 8), rep("Thetayc", 2), rep(c("Sobs", "Shannon", "Evenness"), 2))
comparison <- c(rep("Chemotherapy", 4), rep("Radiation", 4), "Chemotherapy", "Radiation", 
                rep("Chemotherapy", 3), rep("Radiation", 3))

st3 <- cbind(variable, comparison, format(chemo_rad_summary$pvalue, digits = 4), format(chemo_rad_summary$bh, digits = 4))

colnames(st3) <- c("Variable Tested", "Treatment", "P-value", "BH Corrected P-value")

@@

>>>>

\newpage
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

\blandscape

@@

variable <- c(rep("sobs", 2), rep("Shannon", 2), rep("Evenness", 2), rep("Thetayc", 2), rep("Positive Probability", 4))
comparison <- rep(c("Chemotherapy", "Radiation"), 6)

st3 <- cbind(variable, comparison, format(chemo_rad_summary$chemo_rad_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$removal_only_mean, digits = 1, nsmall = 1), format(chemo_rad_summary$cr_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$ro_sd, digits = 1, nsmall = 1), format(chemo_rad_summary$pvalue, scientific = TRUE, digits = 2), format(chemo_rad_summary$bh, scientific = TRUE, digits = 2))

colnames(st3) <- c("Variable", "Additive Treatment", "Additive Treatment Mean", "Removal Mean", "Additive Treatment SD", "Removal SD", "P-value", "BH Corrected")

@@

\elandscape
